Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-06-13 paper-new/concrete-term-model.tex
[paper] work on concrete term model section

In this section, we use SGDT to give a set-theoretic denotational semantics to
the terms of the gradually-typed lambda calculus introduced in Section
\ref{sec:GTLC}. This serves two purposes: First, it is a simple setting in which
to employ the tools of SGDT. Second, constructing this semantic model
establishes the validity of the beta and eta principles for the gradually-typed
lambda calculus.
%
In Section \ref{sec:towards-relational-model}, we will discuss how to extend the
denotational semantics to accommodate the type and term precision orderings.

@@

We begin by discussing the objects of the model into which we will embed
the types and terms of the gradually-typed lambda calculus. Recall that we are
working in an ambient guarded type theory (see Section
\ref{sec:technical-background}) for details. The constructions in this section
will make significant use of the primitives of guarded type theory.

@@

We now describe the monad that decomposes into the free-forgetful adjunction
between sets and error domains. We base the construction on the \emph{guarded
lift monad} described in \cite{mogelberg-paviotti2016}. Here, we augment the
guarded lift monad to accommodate the additional effect of failure. For a type
$A$, we define the \emph{guarded lift monad with failure} $\li A$, which we will
just call the \emph{guarded lift monad} or simply the \emph{lift monad}.

\begin{align*}
  \li A &:= \\
  &\eta \colon A \to \li A \\
  &\mho \colon \li A \\
  &\theta \colon \laterhs (\li A) \to \li A
\end{align*}

Unless otherwise mentioned, all constructs involving $\later$ or $\fix$ are
understood to be with respect to a fixed clock $k$. So for the above, we really
have for each clock $k$ a type $\li^k A$ with respect to that clock.
%
Formally, the lift monad $\li A$ is defined as the solution to the guarded
recursive type equation
%
\[ \li A \cong A + 1\, + \laterhs \li A. \]
%
An element of $\li A$ should be viewed as a computation that can either (1)
return a value (via $\eta$), (2) raise an error and stop (via $\mho$), or (3)
take a step (via $\theta$).
%
Notice there is a computation $\fix\, \theta$ of type $\li A$. This represents a
computation that runs forever and never returns a value.

@@

where the $\li D$ represents the fact that in the case where the value of type
$\dyn$ is a function, the function when called may return an error or take a
step of computation. Unfortunately, this equation does not have inductive or
coinductive solutions. The usual way of dealing with such equations is via
domain theory, by which we can obtain an exact solution. However, the machinery
of domain theory can be difficult for language designers to learn and apply in
the mechanized setting. Instead, we will leverage the tools of guarded type
theory. Consider the following similar-looking equation:
%
\[ D \cong \Nat + (D \times D)\, + \laterhs (D \to \li D). \]
%
Since the negative occurrence of $D$ is guarded under a later, this equation has
a solution constructed via a combination of least fixed-point and guarded
recursion. Specifically, consider the parameterized inductive type
%
\[ D'\, X := \mu T. \Nat + (T \times T) + X. \]
%
Now consider the function $f$ defined by
%
\[ \lambda (\tilde{D} \colon \laterhs \type) . D' (\later_t (\tilde{D}_t \to (\li (\tilde{D}_t)))). \]
%
Recall that the tick $t : \tick$ is evidence that time has passed, and since
$\tilde{D}$ has type $\later \type$, i.e. $\tick \to \type$, then $\tilde{D}_t$
has type $\type$.

Now, define 
\[ D := \fix f. \]
%
% TODO refer to technical background?
Observe that the unfolding equality of the guarded fixpoint implies that $D = D'
(\later D \to \li D)$, and by unfolding $D'$ we see that this is equal to $\Nat
+ (D \times D) + \later (D \to \li D)$.

@@

We can now give a semantics to the terms of the gradually-typed lambda calculus.
The full definition is given in Figure \ref{fig:term-semantics}.
%
Much of the semantics is similar to a normal call-by-value denotational
semantics: We interpret types as sets, and terms as functions. Contexts $\Gamma
= x_1 \colon A_1, \dots, x_n \colon A_n$ is interpreted as the product
$\sem{A_1} \times \cdots \times \sem{A_n}$. The semantics of the dynamic type
$\dyn$ is the set $D$ introduced in Section \ref{sec:dynamic-type}. The product
type $A \times A'$ is interpreted as the product of the denotations of $A$ and
$A'$. The function type $A \ra A'$ is interpreted as the set of functions from
$\sem{A}$ to $\li (\sem {A'})$.

@@

The semantics of the casts for a type precision derivation $c$ are defined by
induction on $c$. The interesting cases are $\inat$, $\itimes$, and $\iarr$.
Recall that $\Dyn$ is isomorphic to $\Nat\, + (\Dyn \times \Dyn) + \laterhs
(\Dyn \to \li \Dyn)$ (say that the sum is left-associative).
%
For $\inat$, the upcast is simply $\inl \circ \inl$. The downcast has type $\Dyn
\to \li \Nat$ and performs a case inspection on the sum type. If it's a natural
number $n$, then we return $\eta\, n$. Otherwise, we return $\mho$, modelling
the fact that a run-time error occurs.
%
For $\itimes$, the upcast is $\inr \circ \inl$, and the downcast performs a case
inspection analogously to the natural number downcast.
%
% TODO explain this further
For $\iarr$, the upcast is given by $\inr \circ \nxt$. The downcast performs a
case inspection, and in the $\inr$ case, it uses a $\theta$ in order to gain
access to the function under the $\later$.
%
% TODO write this out explicitly?

% TODO function casts

@@

\vspace{2ex}

\begin{figure*}
  \noindent Types:
  \begin{align*}
    \sem{\nat} &= \Nat \\
    \sem{\dyn} &= \Dyn \\
    \sem{A \times A'} &= \sem{A} \times \sem{A'} \\ 
    \sem{A \ra A'} &= \sem{A} \To \li \sem{A'} \\
  \end{align*}

@@

  % TODO check these
  \noindent Values and terms:
  \begin{align*}
    \sem{\zro}         &= \lambda \gamma . 0 \\
    \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
    \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
    \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
    \sem{V_f\, V_x}    &= \lambda \gamma . {({(\sem{V_f}\, \gamma)} \, {(\sem{V_x}\, \gamma)})} \\
    \sem{\err_B}       &= \lambda \gamma . \mho \\
    %
    % upcasts
    \sem{\upc{\inat{V_n}}} &= \lambda \gamma . \inl\, \inl\, (\sem{V_n}\, \gamma) \\
    \sem{\upc{\itimes{V_1, V_2}}} &= \lambda \gamma . \inl\, \inr\, (\sem{V_1}\, \gamma, \sem{V_2}\, \gamma) \\
    \sem{\upc{\iarr{V_f}}} &= \lambda \gamma . \inr\, \nxt (\sem{V_f}\, \gamma) \\[2ex]
    %
    % downcasts
    \sem{\dnc{\inat V_d}} &=
      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
      \{ \inl\, (\inl\, n) \to \eta\, n
         \alt \text{otherwise} \to \mho \} \\
    \sem{\dnc{\itimes V_d}} &=
      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
      \{ \inl\, (\inr\, (d_1, d_2)) \to \eta\, (d_1, d_2)
         \alt \text{otherwise} \to \mho \} \\
    \sem{\dnc{\iarr V_d}} &=
      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
      \{ \inr\, \tilde{f} \to \theta\, (\lambda t. \eta (\tilde{f}_t))
         \alt \text{otherwise} \to \mho \} \\
    %
    % TODO product and function
    %
    % \sem{\casedyn{V}{n} {M_{nat}} {\tilde{f}}{M_{fun}}} &=
    % \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
    % &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{nat}}(n) \\
    % &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{fun}}(\tilde{f})
    %
    %
    % \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
    % \sem{\bind{x}{M}{N}} &= \lambda \delta . \ext {(\lambda x . \sem{N}\, (\delta, x))} {\sem{M}\, \delta} \\
  \end{align*}

@@

>>>>

\eric{Wording?}
The principal goal of the next two sections is to ground the refined notions of
abstract model of gradual typing described in Section \ref{sec:abstract-models}
in a concrete setting. We aim to introduce the key issues whose resolutions in
the concrete setting inspire the refined definitions of abstract model.

In the present section, we use the techniques of SGDT to give a set-theoretic
denotational semantics to the terms of the gradually-typed lambda calculus
introduced in Section \ref{sec:GTLC}. This serves two important purposes: First,
it acts as a warm-up for employing the tools of SGDT to define a denotational
semantics. Second, constructing this semantic model establishes the validity of
the $\beta$ and $\eta$ principles for the gradually-typed lambda calculus. It
also acts as a stepping stone into the next section, where we extend the
techniques employed here to accommodate the type and term precision orderings of
the gradually-typed lambda calculus.
%
% In Section \ref{sec:towards-relational-model}, we will discuss how to extend the
% denotational semantics to accommodate the type and term precision orderings.

%necessitate the abstract components that are found in the definition of the model.

@@

We begin by discussing the objects of the model into which we will embed the
types and terms of the gradually-typed lambda calculus. Recall that we are
working informally in a guarded type theory with clocks and ticks (see Section
\ref{sec:technical-background}) for details. The constructions in this section
will make significant use of the primitives of guarded type theory.

@@

We now describe the monad that decomposes into the free-forgetful adjunction
between sets and error domains. We base the construction on the \emph{guarded
lift monad} described in \cite{mogelberg-paviotti2016}. Here, we augment the
guarded lift monad to accommodate the additional effect of failure. For a type
$A$, we define the \emph{guarded lift monad with failure} $\li A$, which we will
just call the \emph{guarded lift monad} or simply the \emph{lift monad}.
%
\begin{align*}
  \li A :=\quad
  &\alt \eta \colon A \to \li A \\
  &\alt \mho \colon \li A \\
  &\alt \theta \colon \laterhs (\li A) \to \li A
\end{align*}
%
Unless otherwise mentioned, all constructs involving $\later$ or $\fix$ are
understood to be with respect to a fixed clock $k$. So for the above, we really
have for each clock $k$ a type $\li^k A$ with respect to that clock.
%
Formally, the lift monad $\li A$ is defined as the solution to the guarded
recursive type equation
%
\[ \li A \cong A + 1\, + \laterhs \li A. \]
%
An element of $\li A$ should be viewed as a computation that can either (1)
return a value (via $\eta$), (2) raise an error and stop (via $\mho$), or (3)
take a step (via $\theta$).
%
Notice there is a computation $\fix\, \theta$ of type $\li A$. This represents a
computation that runs forever and never returns a value.

@@

where the $\li D$ represents the fact that in the case where the value of type
$\dyn$ is a function, the function when called may return an error or take a
step of computation. Unfortunately, this equation does not have inductive or
coinductive solutions. The usual way of dealing with such equations is via
domain theory, by which we can obtain an exact solution. However, the machinery
of domain theory can be difficult for language designers to learn and apply in
the mechanized setting. Instead, we will leverage the tools of guarded type
theory. Consider the following similar-looking equation:
%
\[ D \cong \Nat + (D \times D)\, + \laterhs (D \to \li D). \]
%
Since the negative occurrence of $D$ is guarded under a later, this equation has
a solution constructed via a combination of least fixed-point and guarded
recursion. Specifically, consider the parameterized inductive type
%
\[ D'\, X := \mu T. \Nat + (T \times T) + X. \]
%
Now consider the function $f$ defined by
%
\[ \lambda (\tilde{D} \colon \laterhs \type) . D' (\laterhs_t (\tilde{D}_t \to (\li (\tilde{D}_t)))). \]
%
Recall that the tick $t : \tick$ is evidence that time has passed, and since
$\tilde{D}$ has type $\later \type$, i.e. $\tick \to \type$, then $\tilde{D}_t$
has type $\type$.
%
Finally, define 
\[ D := \fix f. \]
%
% TODO refer to technical background?
Observe that the unfolding equality of the guarded fixpoint implies that $D = D'
(\later D \to \li D)$, and by the definition of $D'$ as a least fixpoint we see
that this is equal to $\Nat + (D \times D) + \later (D \to \li D)$.

@@

We can now give a semantics to the terms of the gradually-typed lambda calculus.
The definition of the interpretation of types and cast terms is given in Figure
\ref{fig:term-semantics}.
%
Much of the semantics is similar to a normal call-by-value denotational
semantics: We interpret types as sets, and terms as functions. Contexts $\Gamma
= x_1 \colon A_1, \dots, x_n \colon A_n$ are interpreted as the product
$\sem{A_1} \times \cdots \times \sem{A_n}$. The semantics of the dynamic type
$\dyn$ is the set $D$ introduced in Section \ref{sec:dynamic-type}. The product
type $A \times A'$ is interpreted as the product of the denotations of $A$ and
$A'$. The function type $A \ra A'$ is interpreted as the set of functions from
$\sem{A}$ to $\li (\sem {A'})$.

@@

The semantics of the casts for a type precision derivation $c$ are defined by
induction on $c$. The interesting cases are $\inat$, $\itimes$, and $\iarr$.
Recall that $D$ is isomorphic to $\Nat\, + (D \times D)\, + \laterhs (D \to \li
D)$ (say that the sum is left-associative).
%
For $\inat$, the upcast is simply $\inl \circ \inl$. The downcast has type $D
\to \li D$ and performs a case inspection on the sum type. If it is a natural
number $n$, then we return $\eta\, n$. Otherwise, we return $\mho$, modelling
the fact that a run-time error occurs.
%
For $\itimes$, the upcast is $\inr \circ \inl$, and the downcast performs a case
inspection analogously to the natural number downcast.
%
% TODO explain this further
For $\iarr$, the upcast is given by $\inr \circ \nxt$. The downcast performs a
case inspection, and in the $\inr$ case, it uses a $\theta$ in order to gain
access to the function under the $\later$.
%
% TODO write this out explicitly?

% TODO should we include function casts?

@@

\begin{figure*}
  \noindent Types:
  \begin{align*}
    \sem{\nat} &= \Nat \\
    \sem{\dyn} &= D \\
    \sem{A \times A'} &= \sem{A} \times \sem{A'} \\ 
    \sem{A \ra A'} &= \sem{A} \To \li \sem{A'} \\
  \end{align*}

@@

  % TODO check these
  \noindent Casts:
  \begin{align*}
    % \sem{\zro}         &= \lambda \gamma . 0 \\
    % \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
    % \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
    % \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
    % \sem{V_f\, V_x}    &= \lambda \gamma . {({(\sem{V_f}\, \gamma)} \, {(\sem{V_x}\, \gamma)})} \\
    % \sem{\err_B}       &= \lambda \gamma . \mho \\
    %
    % upcasts
    \sem{\upc{\inat{V_n}}} &= \lambda \gamma . \inl\, \inl\, (\sem{V_n}\, \gamma) \\
    \sem{\upc{\itimes{V_1, V_2}}} &= \lambda \gamma . \inl\, \inr\, (\sem{V_1}\, \gamma, \sem{V_2}\, \gamma) \\
    \sem{\upc{\iarr{V_f}}} &= \lambda \gamma . \inr\, \nxt (\sem{V_f}\, \gamma) \\[2ex]
    %
    % downcasts
    \sem{\dnc{\inat V_d}} &=
      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
      \{ \inl\, (\inl\, n) \to \eta\, n
         \alt \text{otherwise} \to \mho \} \\
    \sem{\dnc{\itimes V_d}} &=
      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
      \{ \inl\, (\inr\, (d_1, d_2)) \to \eta\, (d_1, d_2)
         \alt \text{otherwise} \to \mho \} \\
    \sem{\dnc{\iarr V_d}} &=
      \lambda \delta . \text{case $(\sem{V_d}\, \gamma)$ of}
      \{ \inr\, \tilde{f} \to \theta\, (\lambda t. \eta (\tilde{f}_t))
         \alt \text{otherwise} \to \mho \} \\
    %
    % TODO product and function
    %
    % \sem{\casedyn{V}{n} {M_{nat}} {\tilde{f}}{M_{fun}}} &=
    % \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
    % &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{nat}}(n) \\
    % &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{fun}}(\tilde{f})
    %
    %
    % \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
    % \sem{\bind{x}{M}{N}} &= \lambda \delta . \ext {(\lambda x . \sem{N}\, (\delta, x))} {\sem{M}\, \delta} \\
  \end{align*}

@@

\eric{TODO: big-step semantics via clock quantification}

\subsection{A Big-Step Term Semantics}
